### 2024-03-28 doc/15yrsChange.Rmd
updating rmarkdown citations

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes (Lovett et al. 2016, Fischer et al. 2013, Herms and McCullough 2014, Fei et al. 2014). Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement (Fischer et al. 2013, Aukema et al. 2010, Brockerhoff et al. 2006). Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species (Aukema et. al 2010,Anderson-Teixeria 2021, Klooster et al. 2013, Ellison et al. 2018). Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al. 2013).

Another category of nuisance species is mammals whose population densities have been inflated due to human activities (Murphey et al. 2013, Clout & Russel), including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years (McShea et al. 1997), due in large part to global change (McGarvey et al. 2013, Brown et al. 2000, Côté et al. 2004, Rooney 2001). White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density (McGarvey et al. 2013). Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species (Bugalho et al. 2013) while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory (McGarvey et al. 2013; Russell et al., 2017; Miller & McGill, 2019; Gorchov et al., 2021; Miller et al., 2023).

In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.  When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future  (Anderson-Teixeira et al., 2013; McDowell et al., 2020, CITATION). In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die (Miller & McGill, 2019; Miller et al., 2023). The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., Miller & McGill (2019); Nowacki & Abrams (2015)], but even these are often not present in sufficient abundance to replace the current canopy (??).

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al. 2013). The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world (Anderson-Teixeira et al. 2015, Davies et al. 2021).

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm et al. 2013). The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by McGregor et al. (2021). We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) (Anderson-Teixeira et al. 2021).

>>>>

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @fischer_disturbances_2013; @herms_emerald_2014; @fei_biogeomorphic_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021a; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years (McShea et al. 1997), due in large part to global change (McGarvey et al. 2013, Brown et al. 2000, Côté et al. 2004, Rooney 2001). White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.  When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp.,[@miller_compounding_2019; @nowacki_climate_2015]), but even these are often not present in sufficient abundance to replace the current canopy (??).

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [SI archives, @bourg_initial_2013. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021a]
